Federated Learning's Blessing:\\ FedAvg has Linear Speedup

### 2020-06-03 source/sec_overparameterized.tex
minor updates overparameterized

assured~\cite{allen2018convergence,zhang2016understanding}.  Due to the

@@

and we show that it preserves the linear speedup at the same time.  

@@

improved convergence rate over FedAvg. The proof is deferred to Appendix Section~\ref{sec:interpolation}.

@@

\subsection{Geometric Convergence of FedAvg in the Overparameterized Setting}
Recall the federated learning problem $\min_{w}\sum_{k=1}^{N}p_{k}F_{k}(\mathbf{w})$
with $F_{k}(\mathbf{w})=\frac{1}{n_{k}}\sum_{j=1}^{n_{k}}\ell(\mathbf{w};\mathbf{x}_{k}^{j})$.
In this section, we consider the standard Empirical Risk Minimization (ERM) setting where $\ell$
is non-negative, $l$-smooth, and convex, and as before, we assume
that each $F_{k}(\mathbf{w})$ is $L$-smooth and $\mu$-strongly convex. Note that $l\geq L$. This
setup includes many important problems in practice. In the overparameterized
setting, there exists $\mathbf{w}^{\ast}\in\arg\min_{w}\sum_{k=1}^{N}p_{k}F_{k}(\mathbf{w})$
such that $\ell(\mathbf{w}^{\ast};\mathbf{x}_{k}^{j})=0$ for all
$\mathbf{x}_{k}^{j}$. We first show that FedAvg with local SGD updates
and communication every $E$ iterations achieves geometric convergence
with linear speedup in the number of workers. 
\begin{theorem}
	\label{thm:overparameterized_general}In the overparameterized setting,
	FedAvg with local SGD updates and communication every $E$ iterations
	with constant step size $\overline{\alpha}=\mathcal{O}(\frac{1}{E}\frac{N}{l\nu_{\max}+L(N-\nu_{\min})})$
	has geometric convergence,
	\begin{align*}
	\mathbb{E}F(\overline{\mathbf{w}}_{T}) & \leq\frac{L}{2}(1-\overline{\alpha})^{T}\|\mathbf{w}_{0}-\mathbf{w}^{\ast}\|^{2}=\mathcal{O}\left(L\exp(-\frac{\mu}{E}\frac{NT}{l\nu_{\max}+L(N-\nu_{\min})})\cdot\|\mathbf{w}_{0}-\mathbf{w}^{\ast}\|^{2}\right).
	\end{align*}
\end{theorem}
%
\textbf{Linear speedup and Communication Complexity} When $l>L$,
the speedup factor is on the order of $\mathcal{O}(N/E)$ %\frac{N}{E(l\nu_{\max}+L(N-\nu_{\min}))}/\frac{1}{l\nu_{\max}+L(1-\nu_{\min})}\approx\frac{Nl}{E(l+L(N-1))}=
for $N\leq\mathcal{O}(\frac{l}{L})$, i.e. FedAvg with $N$ workers and communication
every $E$ iterations provides a geometric convergence speedup factor
of $\mathcal{O}(N/E)$, for $N\leq\mathcal{O}(\frac{l}{L})$. When $N$ is above
this threshold, however, the speedup is almost constant in the number
of workers. This matches the findings in \cite{ma2017power}. Our
result also illustrates that $E$ can be taken $\mathcal{O}(T^{\beta})$
for any $\beta<1$, achieving better communication efficiency than
the standard FL setting.

@@

The local device objectives are now given by the sum of squares {\small$F_{k}(\mathbf{w})=\frac{1}{2n_{k}}\sum_{j=1}^{n_{k}}(\mathbf{w}^{T}\mathbf{x}_{k}^{j}-z_{k}^{j})^{2}$},
and there exists $\mathbf{w}^{\ast}$ such that $F(\mathbf{w}^{\ast})\equiv0$.
Define the local Hessian matrix as $H^{k}:=\frac{1}{n_{k}}\sum_{j=1}^{n_{k}}\mathbf{x}_{k}^{j}(\mathbf{x}_{k}^{j})^{T}$,
the averaged Hessian matrix as $H:=\sum_{k=1}^{N}p_{k}H^{k}$, and the stochastic Hessian matrix as $\tilde{H}_{t}^{k}:=\xi_{t}^{k}(\xi_{t}^{k})^{T}$. 
\begin{comment}
In general $H$ has zero eigevalues. However, because the null space
of $H$ and range of $H$ are orthogonal, in our subsequence analysis
it suffices to project $\overline{\mathbf{w}}_{t}-\mathbf{w}^{\ast}$
onto the range of $H$, thus we may restrict to the non-zero eigenvalue
of $H$. We can use $\mathbf{w}^{\ast T}\mathbf{x}_{k}^{j}-z_{k}^{j}\equiv0$
to rewrite the local objectives as $F_{k}(\mathbf{w})=\frac{1}{2}\langle\mathbf{w}-\mathbf{w}^{\ast},H^{k}(\mathbf{w}-\mathbf{w}^{\ast})\rangle\equiv\frac{1}{2}\|\mathbf{w}-\mathbf{w}^{\ast}\|_{H^{k}}^{2}$
so that $F(\mathbf{w})=\frac{1}{2}\|\mathbf{w}-\mathbf{w}^{\ast}\|_{H}^{2}$.
\end{comment}
\begin{comment}
\begin{align*}
F_{k}(w) & =\frac{1}{2n_{k}}\sum_{j=1}^{n_{k}}(w^{T}x_{k,j}-z_{k,j}-(w^{\ast T}x_{k,j}-z_{k,j}))^{2}=\frac{1}{2n_{k}}\sum_{j=1}^{n_{k}}((w-w^{\ast})^{T}x_{k,j})^{2}\\
& =\frac{1}{2}\langle w-w^{\ast},H^{k}(w-w^{\ast})\rangle=\frac{1}{2}\|w-w^{\ast}\|_{H^{k}}^{2}
\end{align*}
\end{comment}
\begin{comment}
Let $\xi_{t}^{k}$ be the stochastic sample on the $k$th device at
time $t$, and define $\tilde{H}_{t}^{k}:=\xi_{t}^{k}(\xi_{t}^{k})^{T}$
as the stochastic Hessian matrix. %	
\end{comment}
\begin{comment}
Note that $\mathbb{E}\tilde{H}_{t}^{k}=\frac{1}{n_{k}}\sum_{j=1}^{n_{k}}\mathbf{x}_{k}^{j}(\mathbf{x}_{k}^{j})^{T}=H^{k}$
and $\mathbf{g}_{t,k}=\nabla F_{k}(\mathbf{w}_{t}^{k},\xi_{t}^{k})=\tilde{H}_{t}^{k}(\mathbf{w}_{t}^{k}-\mathbf{w}^{\ast})$
while $\mathbf{g}_{t}=\sum_{k=1}^{N}p_{k}\nabla F_{k}(\mathbf{w}_{t}^{k},\xi_{t}^{k})=\sum_{k=1}^{N}p_{k}\tilde{H}_{t}^{k}(\mathbf{w}_{t}^{k}-\mathbf{w}^{\ast})$.
\end{comment}
Define $l$ to be the smallest positive number such that $\mathbb{E}\|\xi_{t}^{k}\|^{2}$$\mathbf{\xi}_{t}^{k}$($\mathbf{\xi}_{t}^{k})^{T}\preceq lH^{k}$
for all $k$. Note that $l\leq\max_{k,j}\|\mathbf{x}_{k}^{j}\|^{2}$.
Let $L$ and $\mu$ be lower and upper bounds of non-zero eigenvalues
of $H^{k}$. Define $\kappa_{1}:=l/\mu$ and $\kappa:=L/\mu$. Following
\cite{liu2018accelerating,jain2017accelerating}, we define the statistical
condition number $\tilde{\kappa}$ as the smallest positive real number
such that $\mathbb{E}\left[\langle\xi_{t}^{k}(H^{k})^{-1},\xi_{t}^{k}\rangle\xi_{t}^{k}(\xi_{t}^{k})^{T}\right] \  \preceq\tilde{\kappa}H^{k}$, for all $k$. 
The condition numbers $\kappa_{1}$ and $\tilde{\kappa}$
are important in the characterization of convergence rates for FedAvg
algorithms. Note that $\kappa_{1}>\kappa$ and $\kappa_{1}>\tilde{\kappa}$.

@@

\begin{theorem}
	\label{thm:overparameterized_quadratic}For the overparamterized quadratic
	problem, FedAvg with local SGD updates and communication every $E$
	iterations with constant step size $\overline{\alpha}=\mathcal{O}(\frac{1}{E}\frac{N}{l\nu_{\max}+\mu(N-\nu_{\min})})$
	has geometric convergence:
	\begin{align*}
	\mathbb{E}F(\overline{\mathbf{w}}_{T}) & \leq\mathcal{O}\left(L(1-\frac{N}{E(\nu_{\max}\kappa_{1}+(N-\nu_{\min}))})^{T}\|\mathbf{w}_{0}-\mathbf{w}^{\ast}\|^{2}\right).
	\end{align*}
\end{theorem}

>>>>

ensured~\cite{allen2018convergence,zhang2016understanding}.  Due to the

@@

and show that it preserves linear speedup at the same time.  

@@

improved convergence rate over FedAvg. Detailed proofs are deferred to Appendix Section~\ref{sec:interpolation}.

@@

\subsection{Geometric Convergence of FedAvg in the Overparameterized Setting}
Recall the federated learning problem $\min_{w}\sum_{k=1}^{N}p_{k}F_{k}(\mathbf{w})$
with $F_{k}(\mathbf{w})=\frac{1}{n_{k}}\sum_{j=1}^{n_{k}}\ell(\mathbf{w};\mathbf{x}_{k}^{j})$.
In this section, we consider the standard Empirical Risk Minimization (ERM) setting where $\ell$
is non-negative, $l$-smooth, and convex, and as before, each $F_{k}(\mathbf{w})$ is $L$-smooth and $\mu$-strongly convex. Note that $l\geq L$. This
setup includes many important problems in practice. In the overparameterized
setting, there exists $\mathbf{w}^{\ast}\in\arg\min_{w}\sum_{k=1}^{N}p_{k}F_{k}(\mathbf{w})$
such that $\ell(\mathbf{w}^{\ast};\mathbf{x}_{k}^{j})=0$ for all
$\mathbf{x}_{k}^{j}$. We first show that FedAvg achieves geometric convergence
with linear speedup in the number of workers. 
\begin{theorem}
	\label{thm:overparameterized_general}In the overparameterized setting,
	FedAvg with communication every $E$ iterations
	and constant step size $\overline{\alpha}=\mathcal{O}(\frac{1}{E}\frac{N}{l\nu_{\max}+L(N-\nu_{\min})})$
	has geometric convergence
	\begin{align*}
	\mathbb{E}F(\overline{\mathbf{w}}_{T}) & \leq\frac{L}{2}(1-\overline{\alpha})^{T}\|\mathbf{w}_{0}-\mathbf{w}^{\ast}\|^{2}=\mathcal{O}\left(L\exp(-\frac{\mu}{E}\frac{NT}{l\nu_{\max}+L(N-\nu_{\min})})\cdot\|\mathbf{w}_{0}-\mathbf{w}^{\ast}\|^{2}\right).
	\end{align*}
\end{theorem}
%
\textbf{Linear speedup and Communication Complexity} The linear speedup factor is on the order of $\mathcal{O}(N/E)$ %\frac{N}{E(l\nu_{\max}+L(N-\nu_{\min}))}/\frac{1}{l\nu_{\max}+L(1-\nu_{\min})}\approx\frac{Nl}{E(l+L(N-1))}=
for $N\leq\mathcal{O}(\frac{l}{L})$, i.e. FedAvg with $N$ workers and communication
every $E$ iterations provides a geometric convergence speedup factor
of $\mathcal{O}(N/E)$, for $N\leq\mathcal{O}(\frac{l}{L})$. When $N$ is above
this threshold, however, the speedup is almost constant in the number
of workers. This matches the findings in \cite{ma2017power}. Our
result also illustrates that $E$ can be taken $\mathcal{O}(T^{\beta})$
for any $\beta<1$ to achieve geometric convergence, achieving better communication efficiency than
the standard FL setting.

@@

The local device objectives are now given by the sum of squares {\small$F_{k}(\mathbf{w})=\frac{1}{2n_{k}}\sum_{j=1}^{n_{k}}(\mathbf{w}^{T}\mathbf{x}_{k}^{j}-z_{k}^{j})^{2}$},
and there exists $\mathbf{w}^{\ast}$ such that $F(\mathbf{w}^{\ast})\equiv0$.
Define the local Hessian matrix as $H^{k}:=\frac{1}{n_{k}}\sum_{j=1}^{n_{k}}\mathbf{x}_{k}^{j}(\mathbf{x}_{k}^{j})^{T}$, and the stochastic Hessian matrix as $\tilde{H}_{t}^{k}:=\xi_{t}^{k}(\xi_{t}^{k})^{T}$. 
\begin{comment}
In general $H$ has zero eigevalues. However, because the null space
of $H$ and range of $H$ are orthogonal, in our subsequence analysis
it suffices to project $\overline{\mathbf{w}}_{t}-\mathbf{w}^{\ast}$
onto the range of $H$, thus we may restrict to the non-zero eigenvalue
of $H$. We can use $\mathbf{w}^{\ast T}\mathbf{x}_{k}^{j}-z_{k}^{j}\equiv0$
to rewrite the local objectives as $F_{k}(\mathbf{w})=\frac{1}{2}\langle\mathbf{w}-\mathbf{w}^{\ast},H^{k}(\mathbf{w}-\mathbf{w}^{\ast})\rangle\equiv\frac{1}{2}\|\mathbf{w}-\mathbf{w}^{\ast}\|_{H^{k}}^{2}$
so that $F(\mathbf{w})=\frac{1}{2}\|\mathbf{w}-\mathbf{w}^{\ast}\|_{H}^{2}$.
\end{comment}
\begin{comment}
\begin{align*}
F_{k}(w) & =\frac{1}{2n_{k}}\sum_{j=1}^{n_{k}}(w^{T}x_{k,j}-z_{k,j}-(w^{\ast T}x_{k,j}-z_{k,j}))^{2}=\frac{1}{2n_{k}}\sum_{j=1}^{n_{k}}((w-w^{\ast})^{T}x_{k,j})^{2}\\
& =\frac{1}{2}\langle w-w^{\ast},H^{k}(w-w^{\ast})\rangle=\frac{1}{2}\|w-w^{\ast}\|_{H^{k}}^{2}
\end{align*}
\end{comment}
\begin{comment}
Let $\xi_{t}^{k}$ be the stochastic sample on the $k$th device at
time $t$, and define $\tilde{H}_{t}^{k}:=\xi_{t}^{k}(\xi_{t}^{k})^{T}$
as the stochastic Hessian matrix. %	
\end{comment}
\begin{comment}
Note that $\mathbb{E}\tilde{H}_{t}^{k}=\frac{1}{n_{k}}\sum_{j=1}^{n_{k}}\mathbf{x}_{k}^{j}(\mathbf{x}_{k}^{j})^{T}=H^{k}$
and $\mathbf{g}_{t,k}=\nabla F_{k}(\mathbf{w}_{t}^{k},\xi_{t}^{k})=\tilde{H}_{t}^{k}(\mathbf{w}_{t}^{k}-\mathbf{w}^{\ast})$
while $\mathbf{g}_{t}=\sum_{k=1}^{N}p_{k}\nabla F_{k}(\mathbf{w}_{t}^{k},\xi_{t}^{k})=\sum_{k=1}^{N}p_{k}\tilde{H}_{t}^{k}(\mathbf{w}_{t}^{k}-\mathbf{w}^{\ast})$.
\end{comment}
Define $l$ to be the smallest positive number such that $\mathbb{E}\|\xi_{t}^{k}\|^{2}$$\mathbf{\xi}_{t}^{k}$($\mathbf{\xi}_{t}^{k})^{T}\preceq lH^{k}$
for all $k$. Note that $l\leq\max_{k,j}\|\mathbf{x}_{k}^{j}\|^{2}$.
Let $L$ and $\mu$ be lower and upper bounds of non-zero eigenvalues
of $H^{k}$. Define $\kappa_{1}:=l/\mu$ and $\kappa:=L/\mu$. Following
\cite{liu2018accelerating,jain2017accelerating}, we define the statistical
condition number $\tilde{\kappa}$ as the smallest positive real number
such that $\mathbb{E}\left[\langle\xi_{t}^{k}(H^{k})^{-1},\xi_{t}^{k}\rangle\xi_{t}^{k}(\xi_{t}^{k})^{T}\right] \  \preceq\tilde{\kappa}H^{k}$, for all $k$. 
The condition numbers $\kappa_{1}$ and $\tilde{\kappa}$
are important in the characterization of convergence rates for FedAvg
algorithms. Note that $\kappa_{1}>\kappa$ and $\kappa_{1}>\tilde{\kappa}$.

@@

\begin{theorem}
	\label{thm:overparameterized_quadratic}For the overparamterized quadratic
	problem, FedAvg with communication every $E$
	iterations with constant step size $\overline{\alpha}=\mathcal{O}(\frac{1}{E}\frac{N}{l\nu_{\max}+\mu(N-\nu_{\min})})$
	has geometric convergence:
	\begin{align*}
	\mathbb{E}F(\overline{\mathbf{w}}_{T}) & \leq\mathcal{O}\left(L(1-\frac{N}{E(\nu_{\max}\kappa_{1}+(N-\nu_{\min}))})^{T}\|\mathbf{w}_{0}-\mathbf{w}^{\ast}\|^{2}\right).
	\end{align*}
\end{theorem}
